|

### 2023-02-09 paper.Rmd
Thu Feb  9 14:00:07 CET 2023: update-remote.sh

<!-- LaTeX Setup -->
\setcounter{page}{1}
<!-- \tableofcontents -->
\thispagestyle{empty}
\newpage
\listoftables
\listoffigures
\thispagestyle{empty}
\newpage

<!--Markdown Text -->

# Abstract {.unnumbered #I}
...

@@

<!--- Describe experiments... -->
...

@@

  * Encoder $g(x)$: Encode input data to a smaller central layer $h$
  * Decoder $d(h)$: Decode and output the the encoded data back to its original dimension

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM model $M_g$ that are used in each DAE-GP generation $g$ for $g\in{1,...,g_{max}}$. 
Instead of initializing and optimizing each model from scratch as done in previous work (e.g. [@dae-gp_2020_rtree] [@dae-gp_2022_symreg]), a separate DAE-LSTM network $\hat{M}$ will be trained on a large population of randomly initialized solutions.  
The trainable parameters $\theta_{\hat{M}}$ obtained after finishing the training procedure of $\hat{M}$ are then used as the starting parameters for each following DAE-LSTM model $M_g$ for $g\in{1,...,g_{max}}$.

@@

<!--- Symbolic regression -->
To test pre-training in DAE-GP this thesis focuses on the domain of real-world symbolic regression problems. Symbolic Regression problems have been one of the first GP applications [@Koza1993GeneticP] and are an actively studied and highly relevant research area. The goal in symbolic regression is to find a mathematical model for a given set of data points [@10.1007/978-3-540-24621-3_22], in real-world symbolic regression these data points are sourced from real-world observations which in contrast to synthetic symbolic regression problem are more likely to contain random noise and bias. Another important challenge in solving real-world symbolic regression problems is the ability for a given model to generalize, we want the final model to show high accuracy in predicting outcomes for previously unseen cases.

<!--- airfoil dataset -->
The main experiments conducted in this thesis uses the NASA Airfoil Self-Noise Data Set [@AIRFOIL_DATASET] that consists of 5 input variables and 1 output variable that are listed in table \@ref(tab:airfoil-dataset-description).

@@

* Variable number of hidden neurons (50, 100, 200) with a single hidden layer
* Fixed number of hidden neurons (100) per hidden layer with variable number of hidden layers (1, 2, 3)

@@

```{r airfoil-popDiversity, echo=FALSE, fig.cap="Median Population Diversity over 30 Generations - Airfoil", out.width="80%"}
knitr::include_graphics("./img/airfoil_2hl_150hn_fullRun_30gens_withGP/median_popDiversity_byGens_levOnly.png")
```
### Dynamic Adjustment of Hidden Neurons

@@

<!-- An interesting observation from studying the distribution of final fitness scores for both algorithms is that pre-trained DAE-GP has a smaller interquartile range and more extreme outliers for high RMSE which explains the slightly worse performance if looking at median best fitness present in figure \@ref(fig:airfoil-median-fitness-30gens).  -->

@@

```{r airfoil-runtime-total, echo=FALSE, fig.cap="Median Runtime - Airfoil", out.width="80%"}
knitr::include_graphics("./img/airfoil_2hl_150hn_fullRun_30gens_TimeBenchmark/median_runtime.png")
```
Figure \@ref(fig:airfoil-runtime-boxplot) shows the distribution of the total runtime for both DAE-GP variants. It shows that Pre-Training, besides largely increasing the median runtime, also leads to a higher dispersion which is shown by both a larger interquartile range as well as by a a standard deviation that increases by a factor of `r round(782.68 / 635.04, 3)` from 635,04 for regular DAE-GP to 782,68 for pre-trained DAE-GP.

@@

```{r airfoil-runtime-noSampling-total, echo=FALSE, fig.cap="Median Runtime excluding Time for Sampling - Airfoil", out.width="80%"}
knitr::include_graphics("./img/airfoil_2hl_150hn_fullRun_30gens_TimeBenchmark/median_runtime_withoutSampling.png")
```
While investigating other reasons for the differences in run-time, I verified that both DAE-GP variants use the exact same number of trainable parameters for each generations DAE-LSTMS models (total number of trainable parameters DAE-GP=`r format(REG_PARAMS, scientific=F)`, Pre-trained DAE-GP=`r format(PT_PARAMS, scientific=F)`). After profiling a single run of DAE-GP and pre-trained DAE-GP using the python `cProfiler` module, my best explanation for the remaining differences in run-time is an inefficiency in my implementation of pre-training. Figure  \@ref(fig:airfoil-profiling) shows the cumulative time spend for the 20 most expensive function calls. Even though it is difficult to identify the exact reasons for the performance hit for pre-trained DAE-GP, it shows two interesting findings: Firstly, both algorithms spend relatively more time during model sampling (calls to `DAE_LSTM.py:208(sample)`) than during model training (calls to `training.py:1303(fit)`). Secondly, it shows that internal error handling of the baseline neural network library `keras` (calls to `traceback_utils.py:138(error_handler)` and `traceback_utils.py:59(error_handler)`) takes up a comparatively larger amount of run time for pre-trained DAE-GP than for regular DAE-GP.

@@

```{r full-run-realWorldSymReg-problems, echo=FALSE, message=FALSE, warning=FALSE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/full_run_realWorldSymReg_problems.csv"),
  caption="Real World Symbolic Regression Benchmark Problems",
  format = "markdown"
)%>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```
To test for statistical differences of the results achieved by either pre-trained or regular DAE-GP, I selected the nonparametric Mann-Whitney-U Test ([@mwu1], [@mwu2]) to test for differences in the distribution underlying the samples sourced from my experiments. Additionally, I included the effect strength of pre-training as measured by Cliffs Delta [@Cliff1993DominanceSO] into the statistical analysis. Better median values (e.g. higher fitness) are printed bold and p-values are marked with asterisk (1,2,3) for $\alpha$ levels ($0.1$, $0.05$, $0.01$). Note that I included the results for the airfoil problem with a single hidden layer into the statistical evaluation which is not visualized during this chapter (see subsection [Reduced Number of Hidden Layers]).

@@

```
Table \@ref(tab:full-run-realWorldSymReg-best-solution-size) shows the median size of the best solution after 30 generations and results for computing the statistical tests. For Airfoil, Boston Housing and Energy Cooling datasets, the median size of the best soultion are equal, only for the concrete dataset a smaller median size for pre-trained DAE-GP was achieved (pre-trained DAE-GP also showed a higher median best fitness for the concrete problem). The results of the Mann-Whitney U Test again show, that differences in the underlying distributions are not statistically significant which also shows in low effect strenghts of using pre-training on the size of the best found solutions (Cliff's Delta ranges from`r min(df_size$Cliffs_Delta)` to `r max(df_size$Cliffs_Delta)`).

@@

The results of the statistical analysis is summarized in table \@ref(tab:full-run-realWorldSymReg-popDiversity).

@@

# Create the data frame from csv
df <- read.csv(
  "/Users/rmn/github/master_thesis/data/summary_table_median_LevDistance.csv",
  sep=",",
  colClasses = c("character", "numeric", "numeric", "numeric", "character", "character")
)

for(i in 1:nrow(df)){
  if(df[i, "DAE.GP"] > df[i, "Pre.Trained_DAE.GP"]){
    df_bold[i, "DAE.GP"] <- sub("^(.*)$", "**\\1**", as.character(round(df[i, "DAE.GP"],2)))
    df_bold[i, "Pre.Trained_DAE.GP"] <- as.character(round(df[i, "Pre.Trained_DAE.GP"],2))
    }
  else if (df[i, "DAE.GP"] < df[i, "Pre.Trained_DAE.GP"]) {
    df_bold[i, "Pre.Trained_DAE.GP"] <- sub("^(.*)$", "**\\1**", as.character(round(df[i, "Pre.Trained_DAE.GP"],2)))
  df_bold[i, "DAE.GP"] <- as.character(round(df[i, "DAE.GP"],2))
  }
}
# shorten column names
col_names <- c("Problem", "Hidden Layers", "DAE-GP", "Pre-Trained", "P-Value", "Cliffs-Delta")

@@

@@

The experiments conducted in the scope of this master thesis did find that the implemented pre-training strategy resulted in an increased population diversity measured by the normalized levenshtein edit distance. Increasing the population diversity can be an important mechanism in controlling the search behavior of population based meta heuristics and is especially useful for increasing the exploration of the search space of possible solutions !QUELLE!.

@@

# Nomenclature {.unnumbered #II}

## Symbols {.unnumbered #IIa}

```{r Symbols, echo=FALSE, message=FALSE, warning=FALSE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/nomenclature.csv", sep = "?"),
  caption = "List of Mathematical Symbols",
  format = "markdown"
)%>% row_spec(0,bold=TRUE )
```
## Abbreviations {.unnumbered #IIb}

@@

# References {.unnumbered #III}

@@

# Statutory Declaration {.unnumbered #IV}

>>>>

<!-- LaTeX Setup -->

\setcounter{page}{1}

<!-- \tableofcontents -->

```{=tex}
\thispagestyle{empty}
\newpage
\listoftables
\listoffigures
\thispagestyle{empty}
\newpage
```
<!--Markdown Text -->

# Abstract {#I .unnumbered}

...

@@

<!--- Describe experiments... -->

...

@@

-   Encoder $g(x)$: Encode input data to a smaller central layer $h$
-   Decoder $d(h)$: Decode and output the the encoded data back to its original dimension

@@

The pre-training strategy used in this thesis is applied to the DAE-LSTM model $M_g$ that are used in each DAE-GP generation $g$ for $g\in{1,...,g_{max}}$.
Instead of initializing and optimizing each model from scratch as done in previous work (e.g. [@dae-gp_2020_rtree] [@dae-gp_2022_symreg]), a separate DAE-LSTM network $\hat{M}$ will be trained on a large population of randomly initialized solutions.\

The trainable parameters $\theta_{\hat{M}}$ obtained after finishing the training procedure of $\hat{M}$ are then used as the starting parameters for each following DAE-LSTM model $M_g$ for $g\in{1,...,g_{max}}$.

@@

<!--- Symbolic regression -->

To test pre-training in DAE-GP this thesis focuses on the domain of real-world symbolic regression problems.
Symbolic Regression problems have been one of the first GP applications [@Koza1993GeneticP] and are an actively studied and highly relevant research area.
The goal in symbolic regression is to find a mathematical model for a given set of data points [@10.1007/978-3-540-24621-3_22], in real-world symbolic regression these data points are sourced from real-world observations which in contrast to synthetic symbolic regression problem are more likely to contain random noise and bias.
Another important challenge in solving real-world symbolic regression problems is the ability for a given model to generalize, we want the final model to show high accuracy in predicting outcomes for previously unseen cases.

<!--- airfoil dataset -->

The main experiments conducted in this thesis uses the NASA Airfoil Self-Noise Data Set [@AIRFOIL_DATASET] that consists of 5 input variables and 1 output variable that are listed in table \@ref(tab:airfoil-dataset-description).

@@

-   Variable number of hidden neurons (50, 100, 200) with a single hidden layer
-   Fixed number of hidden neurons (100) per hidden layer with variable number of hidden layers (1, 2, 3)

@@

```{r airfoil-popDiversity, echo=FALSE, fig.cap="Median Population Diversity over 30 Generations - Airfoil", out.width="80%"}
knitr::include_graphics("./img/airfoil_2hl_150hn_fullRun_30gens_withGP/median_popDiversity_byGens_levOnly.png")
```

### Dynamic Adjustment of Hidden Neurons

@@

<!-- An interesting observation from studying the distribution of final fitness scores for both algorithms is that pre-trained DAE-GP has a smaller interquartile range and more extreme outliers for high RMSE which explains the slightly worse performance if looking at median best fitness present in figure @ref(fig:airfoil-median-fitness-30gens).  -->

@@

```{r airfoil-runtime-total, echo=FALSE, fig.cap="Median Runtime - Airfoil", out.width="80%"}
knitr::include_graphics("./img/airfoil_2hl_150hn_fullRun_30gens_TimeBenchmark/median_runtime.png")
```

Figure \@ref(fig:airfoil-runtime-boxplot) shows the distribution of the total runtime for both DAE-GP variants.
It shows that Pre-Training, besides largely increasing the median runtime, also leads to a higher dispersion which is shown by both a larger interquartile range as well as by a a standard deviation that increases by a factor of `r round(782.68 / 635.04, 3)` from 635,04 for regular DAE-GP to 782,68 for pre-trained DAE-GP.

@@

```{r airfoil-runtime-noSampling-total, echo=FALSE, fig.cap="Median Runtime excluding Time for Sampling - Airfoil", out.width="80%"}
knitr::include_graphics("./img/airfoil_2hl_150hn_fullRun_30gens_TimeBenchmark/median_runtime_withoutSampling.png")
```

While investigating other reasons for the differences in run-time, I verified that both DAE-GP variants use the exact same number of trainable parameters for each generations DAE-LSTMS models (total number of trainable parameters DAE-GP=`r format(REG_PARAMS, scientific=F)`, Pre-trained DAE-GP=`r format(PT_PARAMS, scientific=F)`).
After profiling a single run of DAE-GP and pre-trained DAE-GP using the python `cProfiler` module, my best explanation for the remaining differences in run-time is an inefficiency in my implementation of pre-training.
Figure \@ref(fig:airfoil-profiling) shows the cumulative time spend for the 20 most expensive function calls.
Even though it is difficult to identify the exact reasons for the performance hit for pre-trained DAE-GP, it shows two interesting findings: Firstly, both algorithms spend relatively more time during model sampling (calls to `DAE_LSTM.py:208(sample)`) than during model training (calls to `training.py:1303(fit)`).
Secondly, it shows that internal error handling of the baseline neural network library `keras` (calls to `traceback_utils.py:138(error_handler)` and `traceback_utils.py:59(error_handler)`) takes up a comparatively larger amount of run time for pre-trained DAE-GP than for regular DAE-GP.

@@

```{r full-run-realWorldSymReg-problems, echo=FALSE, message=FALSE, warning=FALSE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/full_run_realWorldSymReg_problems.csv"),
  caption="Real World Symbolic Regression Benchmark Problems",
  format = "markdown"
)%>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

To test for statistical differences of the results achieved by either pre-trained or regular DAE-GP, I selected the nonparametric Mann-Whitney-U Test ([@mwu1], [@mwu2]) to test for differences in the distribution underlying the samples sourced from my experiments.
Additionally, I included the effect strength of pre-training as measured by Cliffs Delta [@Cliff1993DominanceSO] into the statistical analysis.
Better median values (e.g. higher fitness) are printed bold and p-values are marked with asterisk (1,2,3) for $\alpha$ levels ($0.1$, $0.05$, $0.01$).
Note that I included the results for the airfoil problem with a single hidden layer into the statistical evaluation which is not visualized during this chapter (see subsection [Reduced Number of Hidden Layers]).

@@

Table \@ref(tab:full-run-realWorldSymReg-best-solution-size) shows the median size of the best solution after 30 generations and results for computing the statistical tests.
For Airfoil, Boston Housing and Energy Cooling datasets, the median size of the best soultion are equal, only for the concrete dataset a smaller median size for pre-trained DAE-GP was achieved (pre-trained DAE-GP also showed a higher median best fitness for the concrete problem).
The results of the Mann-Whitney U Test again show, that differences in the underlying distributions are not statistically significant which also shows in low effect strenghts of using pre-training on the size of the best found solutions (Cliff's Delta ranges from`r min(df_size$Cliffs_Delta)` to `r max(df_size$Cliffs_Delta)`).

@@

@@

# Create the data frame from csv
df_popDiv <- read.csv(
  "/Users/rmn/github/master_thesis/data/summary_table_median_LevDistance.csv",
  sep=",",
  colClasses = c("character", "numeric", "numeric", "numeric", "character", "character")
)

# Create a new dataframe with the modified values of the two columns,
df_bold <- df_popDiv

for(i in 1:nrow(df_popDiv)){
  if(df_popDiv[i, "DAE.GP"] > df_popDiv[i, "Pre.Trained_DAE.GP"]){
    df_bold[i, "DAE.GP"] <- sub("^(.*)$", "**\\1**", as.character(round(df_popDiv[i, "DAE.GP"],2)))
    df_bold[i, "Pre.Trained_DAE.GP"] <- as.character(round(df_popDiv[i, "Pre.Trained_DAE.GP"],2))
    }
  else if (df_popDiv[i, "DAE.GP"] < df_popDiv[i, "Pre.Trained_DAE.GP"]) {
    df_bold[i, "Pre.Trained_DAE.GP"] <- sub("^(.*)$", "**\\1**", as.character(round(df_popDiv[i, "Pre.Trained_DAE.GP"],2)))
  df_bold[i, "DAE.GP"] <- as.character(round(df_popDiv[i, "DAE.GP"],2))
  }
}
# shorten column names
col_names <- c("Problem", "Hidden Layers", "DAE-GP", "Pre-Trained", "P-Value", "Cliffs-Delta")

@@

The results of the statistical analysis is summarized in table \@ref(tab:full-run-realWorldSymReg-popDiversity).
For all sub experiments, computing the Mann-Whitney U Test returns P-Values that are smaller than 0.01 which indicates a highly statistical significant difference in the distributions.
Looking at the effect strength of pre-training on population diversity, it shows that pre-training does have a strong positive effect (Cliffs Delta ranges from \`r \`)in improving the levenshtein distance between individual solutions inside the population $P_g$

@@

The experiments conducted in the scope of this master thesis did find that the implemented pre-training strategy resulted in an increased population diversity measured by the normalized levenshtein edit distance.
Increasing the population diversity can be an important mechanism in controlling the search behavior of population based meta heuristics and is especially useful for increasing the exploration of the search space of possible solutions !Q
UELLE!.

@@

# Nomenclature {#II .unnumbered}

## Symbols {#IIa .unnumbered}

```{r Symbols, echo=FALSE, message=FALSE, warning=FALSE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/nomenclature.csv", sep = "?"),
  caption = "List of Mathematical Symbols",
  format = "markdown"
)%>% row_spec(0,bold=TRUE )
```

## Abbreviations {#IIb .unnumbered}

@@

# References {#III .unnumbered}

@@

# Statutory Declaration {#IV .unnumbered}
